---
title: "Question 1"
author: "Loi Nguyen"
date: "2024-04-02"
output: html_document
---


## Read data
```{r}
# Read the CSV file
nhl_data <- read.csv("nhl_player_births.csv")

```



## Display distribution of birth month

```{r}
# Assuming 'data' is your data frame with a column 'birth_month'
# Define a vector of month names
month_names <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                 "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Convert the birth_month column in nhl_data to factor with month names as labels
nhl_data$birth_month <- factor(nhl_data$birth_month, levels = 1:12, labels = month_names)

# Now, continue with the rest of your code
# Step 2: Calculate the percentage of players born in each month
birth_month_counts <- table(nhl_data$birth_month) 
# Count the occurrences of each birth month
total_players <- nrow(nhl_data)  # Total number of players

percentages <- (birth_month_counts / total_players) * 100  # Calculate percentages

# Create a data frame with the calculated percentages
df <- data.frame(month = names(percentages), percentage = as.numeric(percentages))

# Define colors for each month
colors <- c("Jan" = "orange", "Feb" = "orange", "Mar" = "orange", "Apr" = "skyblue", 
            "May" = "skyblue", "Jun"  = "skyblue" , "Jul"  = "skyblue" , 
            "Aug" = "skyblue", "Sep" = "skyblue", "Oct" = "skyblue", 
            "Nov" = "skyblue", "Dec" = "skyblue")

# Create the bar plot using ggplot2
# Reorder the months by percentage
df$month <- factor(df$month, levels = names(sort(percentages, decreasing = TRUE)))

# Load the ggplot2 library if not already loaded
library(ggplot2)

# Create the bar plot using ggplot2 with specified colors and ordered months
ggplot(df, aes(x = month, y = percentage/10, fill = month)) +
  geom_bar(stat = "identity") +
  labs(title = "Percentage of NHL Players Born Each Month",
       x = "Birth Month",
       y = "Percentage") +
  scale_fill_manual(values = colors) +  # Use specified colors
  theme_minimal() +  # Apply a minimal theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 0.833, color = "red", linetype = "dashed") +
  annotate("text", x = length(df$month), y = 0.88, label = "Avg: 0.833%", color = "red", size = 3, hjust = 1) +
  scale_y_continuous(labels = scales::percent_format()) +
  guides(fill = FALSE)  # Remove legend
```

## This graph show all the information of players throughout dozens of seasons so let's dive deeper into each season to make sure that nothing interesting is being missed.


## Merging data so it will include only useful columns. 
```{r}
  nhl_rosters <- read.csv("nhl_rosters.csv")
  
  unique_team_codes <- unique(nhl_rosters$team_code)
  nhl_rosters <- read.csv("nhl_rosters.csv")
  nhl_data <- read.csv("nhl_player_births.csv")
  
  # Merge the datasets on player_id
  merged_data <- merge(nhl_rosters, nhl_data, by = "player_id")
  
  # Select and rename columns from the merged dataset
  middle_data <- subset(merged_data, select = c("player_id", "team_code", "season", "birth_month"))
  colnames(middle_data) <- c("player_id", "team_code", "season", "birth_month")
  
  # Print the resulting middle data
  print(middle_data)

```


```{r}
library(dplyr)
result <- middle_data %>%
  group_by(season) %>%
  summarise(total_players = n(),  # Count total players in the season
            first_3_months = sum(birth_month %in% 1:3)) %>%  # Count players born in the first 3 months
  mutate(percent_first_3_months = (first_3_months / total_players) * 100)  # Calculate percentage

# Print the result
print(result)

```



## How dominant the first-quarter-born through the history?
```{r}
# Load the required library
library(ggplot2)


# Plot the line chart
ggplot(result, aes(x = season / 10000, y = percent_first_3_months)) +
  geom_line() +
  geom_point() +
  labs(title = "Percentage of Players Born in First 3 Months by Season",
       x = "Season",
       y = "Percentage of Players") +
  geom_hline(yintercept = 25, color = "red", linetype = "dashed") + 
  annotate("text", x = 2011.5, y = 35, label = "Introduction of\nMalcolm Gladwell’s Outliers", color = "blue", size = 2) +
  geom_segment(aes(x = 2014, y = 34, xend = 2011.5, yend = 30), color = "red", arrow = arrow(length = unit(0.07, "inches")), size = 0.3) +
  annotate("text", x = 1958, y = 35, label = "First NHL\nEntry Draft", color = "blue", size = 2) +
  geom_segment(aes(x = 1958, y = 34, xend = 1962.5, yend = 28), color = "red", arrow = arrow(length = unit(0.1, "inches")), size = 0.3) +
  theme_minimal()

```





## New data for the last chart

```{r}
library(dplyr)

# Assuming your data frame has columns `team_code` and `season`
result <- middle_data %>%
  group_by(season, team_code) %>%
  summarise(total_players = n(),  # Count total players in the season and team_code
            first_3_months = sum(birth_month %in% 1:3)) %>%  # Count players born in the first 3 months
  mutate(percent_first_3_months = (first_3_months / total_players) * 100)  # Calculate percentage

# Print the result
print(result)


```




```{r}
library(dplyr)
library(ggplot2)

library(ggplot2)

# Count the number of seasons for each team
team_season_counts <- result %>%
  group_by(team_code) %>%
  summarize(total_seasons = n_distinct(season))

# Count the number of seasons where first_3_months_players > 25 for each team
team_exceed_25_counts <- result %>%
  filter(percent_first_3_months > 25) %>%
  group_by(team_code) %>%
  summarize(exceed_25_seasons = n_distinct(season))

# Merge the two datasets
team_counts <- merge(team_season_counts, team_exceed_25_counts, by = "team_code", all.x = TRUE)

# Fill NA values with 0 (for teams where no seasons have first_3_months_players > 25)
team_counts[is.na(team_counts)] <- 0

# Print the result
print(team_counts)


```

## Chart to see whether it is dominant in most teams
```{r}

team_counts <- team_counts %>%
  mutate(percent_exceed_25 = exceed_25_seasons / total_seasons)

# Create the bar chart
ggplot(data = team_counts, aes(x = team_code, y = percent_exceed_25)) +
  geom_bar(stat = "identity", fill = "skyblue" , width = 0.5) +
  labs(title = "Proportion of each Team with dominant first_quarter_born players",
       x = "Team",
       y = "Percent") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
   geom_hline(yintercept = 0.25, color = "red", linetype = "dashed") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6)) 

```
